Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-19 ebayes_rnaseq_heterosis.tex
fixed up some of the real data analysis

Independently for each gene, we ran an MCMC conditional on the hyperparameters to obtain empirical Bayes posterior distributions for all gene-specific parameters. Each MCMC was run using 4 chains with initial values randomly drawn by {\tt Stan}. Each chain ran for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons. The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}. If the maximum value of $\hat{R}$ was larger than 1.1, the total length was increased by 500, with the burn-in increased proportionally, and the chain was rerun. This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest. Thus, we calculate estimated effect sizes using the estimated posterior means of the gene-specific parameters using the formula 
\begin{equation}
\mbox{estimated effect size} = \left\{ 
\begin{array}{ll}
\hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
\hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
0 & \mbox{otherwise}.
\end{array} 
\right. 
\label{e:effect_size}
\end{equation}

>>>>

Independently for each gene, we ran an MCMC conditional on the hyperparameters to obtain empirical Bayes posterior distributions for all gene-specific parameters. Each MCMC was run using 4 chains with initial values randomly drawn by {\tt Stan}. Each chain ran for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons. The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}. If the maximum $\hat{R}$ was larger than 1.1 for any parameter, the total number of iterations was increased by 500, with the burn-in increased proportionally, and the chain was rerun. This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest. Thus, we calculate estimated effect sizes using the estimated posterior means of the gene-specific parameters using the formula in equation \ref{e:effect_size} where a 0 indicates an estimated mean expression for the hybrid that is between those of the parents, i.e. not extreme heterosis.
\begin{equation}
\mbox{estimated effect size} = \left\{ 
\begin{array}{ll}
\hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
\hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
0 & \mbox{otherwise}.
\end{array} 
\right. 
\label{e:effect_size}
\end{equation}
